Smart Spreading Factor Assignment for LoRaWANs

### 2019-01-01 paper.tex
simulation env

\begin{equation} \label{eq:bit_rate_sf}
R_{b} = SF * \dfrac{\left[ \dfrac{4}{4+CR} \right] }{ \left[ \dfrac{2^{SF}}{BW} \right]} \ bits/sec
\end{equation}

@@

\par In Figure~\ref{fig:collision}, a LoRaWAN network distributed around a single GW is illustrated. Different color rings represent achievable range of different SF from GW and different color circles represent nodes with selected SF. Most of the end devices which close to the GW will fall into lowest SF (SF7) area section. End devices which close to the GW will probably select lowest SF all the time. Which results a lot of collisions between same SF transmissions. Collision number increases while end device number close to the GW increases.

\begin{figure}
\centering
\includegraphics[width=0.5\textwidth]{collision}
\caption{Collision between nodes close to the GW.}
\label{fig:collision}
\end{figure}

@@

\section{Proposed Technique} \label{Proposed Technique}
\par The collision issue illustrated in Figure~\ref{fig:collision} can be solved if the GW force some of the close end nodes to select higher SF even if they can able to communicate with lower SF. This may result lower collisions due to the orthogonality of different SF as shown in Figure~\ref{fig:collision_solution_single_gw}. Higher SF assigned nodes are drawn with bold circle border in Figure~\ref{fig:collision_solution_single_gw}. However, how GW assign SF to which node becomes an another issue. Increasing a node's SF should be done carefully since higher SF means higher time on air and higher time on air is more prone to collision with other higher SF transmissions. In multiple GW scenarios, this approach may increase the collision number with nodes in other GW's range. Thus, extra care must be taken for nodes in intersection area of GWs illustrated in Figure~\ref{fig:collision_solution_multi_gw}.

@@

\begin{figure}
\centering
\includegraphics[width=0.5\textwidth]{collision_solution_single_gw}
\caption{Collision between nodes close to the GW can be prevented by using higher SF.}
\label{fig:collision_solution_single_gw}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=0.5\textwidth]{collision_solution_multi_gw}
\caption{Collision prevention proposal for intersecting GWs.}
\label{fig:collision_solution_multi_gw}
\end{figure}

\section{Simulation Environment and Results} \label{Simulation Environment and Results}
\par A discrete event simulator is implemented in Python to study LoRa network performance. The LoRaWAN SF simulation tool is publicly available at https://github.com/tugrulyatagan/simlorafs.

Simulation Assumptions

\subsection{Link Model}
\par [TODO]

The link budget of a wireless system or network is a measure of all the gains and losses from the transmitter, through the propagation channel, to the target receiver. Link budget of a wireless link can be calculated as:

@@

\par Where, $P^{dBm}_{RX}$ is the expected receive power at the receiver. $P^{dBm}_{TX}$ is the transmit power. $G^{dB}_{SYS}$ is the system gains such as transmitter and receiver antenna gains. $L^{dB}_{SYS}$ is the system losses such as transmitter and receiver line, circuit, antenna losses. $L^{dB}_{PATH}$ is the propagation path loss between transmitter and receiver antennas.

\par Propagation loss can be calculated as:

@@

\par Where $h$ is the gateway altitude and $f$ is the frequency of the signal. We assume that $h$ = 15 m and $f$ = 868 MHz, Propagation loss can be calculated as:

@@

\par Receive sensitivity of a LoRa gateway for different SFs and BWs can be found in Table \ref{table:gw_sf_sensitivity} \cite{SX1276}. Maximum TX power for European ISM band LoRa default channels can be found in Table \ref{table:max_tx_power}.

\begin{table}
\centering
\caption{Gateway Sensitivity for different SFs}
\label{table:gw_sf_sensitivity}
\begin{tabular}{|c|c|c|c|c|c|c|c|}
\hline
\multicolumn{2}{|c|}{\multirow{2}{*}{}} & \multicolumn{6}{c|}{\textbf{SF}} \\ \cline{3-8}
\multicolumn{2}{|c|}{}                  &    7 &    8 &    9 &   10 &   11 &   12 \\ \hline
\multirow{3}{*}{\textbf{BW}}  & 125 kHz & -123 & -126 & -129 & -132 & -133 & -136 \\ \cline{2-8}
                              & 250 kHz & -120 & -123 & -125 & -128 & -130 & -133 \\ \cline{2-8}
                              & 500 kHz & -116 & -119 & -122 & -125 & -128 & -130 \\ \hline
\end{tabular}
\end{table}

\begin{table}
\centering
\caption{EU863-870 ISM Band LoRa Default Channels}
\label{table:max_tx_power}
\begin{tabular}{|c|c|c|c|}
\hline
\textbf{Channel} & \textbf{Frequency} & \textbf{Bandwidth} & \textbf{Max EIRP} \\ \hline
      1 &     868.1 &   125 kHz &   14 dBm \\ \hline
      2 &     868.3 &   125 kHz &   14 dBm \\ \hline
      3 &     868.5 &   125 kHz &   14 dBm \\

\subsection{Interference Model}
\par [TODO]

@@

\begin{equation} \label{eq:p_interference}
P_{rc,y}^{interf} = \dfrac{P_{rc,y}(T_{x}-t_{1})}{T_{x}}
\end{equation}

\begin{figure}
\centering
\includegraphics[width=0.5\textwidth]{sf_pdr}
\caption{PDR of different SFs.}
\label{fig:sf_pdr}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=0.5\textwidth]{gw_pdr}
\caption{PDR of different gateway numbers.}
\label{fig:gw_pdr}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=0.5\textwidth]{r_pdr}
\caption{PDR of different topology radius.}
\label{fig:r_pdr}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=0.5\textwidth]{pr_pdr}
\caption{PDR of different packet rates.}
\label{fig:pr_pdr}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=0.5\textwidth]{predict_pdr}
\caption{PDR of different prediction SF assignment scheme.}
\label{fig:pr_pdr}
\end{figure}

@@

\section{Conclusion} \label{Conclusion}
\par In this paper, after brief introduction about LPWAN technologies, we provide background information about one of the most promising LPWAN technology, LoRa. Then, we discuss about LoRa modulation basics and spreading factor assignment issue. We have implemented a discrete event simulator from scratch to study network performance of LoRaWAN and evaluate different SF assignment schemes. We have shown how same SF transmission collisions can be avoided and we proposed a machine learning based solution for collision with same SF transmissions. We illustrate simulation results for proposed machine learning SF assignment technique. Our simulation results show that, proposed technique can increase network performance up to [TBD] \%.

\par As for future works, transmit power optimization can be included to proposed machine learning scheme. In this paper, we assume nodes always uses maximum transmit power for uplink transmission, however nodes close to the GW can decrease transmit power to save energy. This will make transmission more vulnerable to interference thus requires extra care. Optimizing another node parameter alongside SF can be interesting research. Also other machine learning methods can be investigated for SF optimization. 
Reinforcement learning can be a good candidate for this network optimization issue.

\par \cite{7815384} \cite{7803607} \cite{7996384} \cite{8090518} \cite{s17061193} \cite{8267219} \cite{8430542} \cite{8319183} \cite{8480649} \cite{AN1200.22} \cite{Bor:2016:LLW:2988287.2989163} \cite{8406255} \cite{DBLP:journals/corr/abs-1802-10338} \cite{finnegan2018comparative}

>>>>

\begin{equation} \label{eq:bit_rate_sf}
R_{b}|_{bps} = SF * \dfrac{\left[ \dfrac{4}{4+CR} \right] }{ \left[ \dfrac{2^{SF}}{BW|_{Hz}} \right]}
\end{equation}

@@

\par In Figure \ref{fig:collision}, a LoRaWAN network distributed around a single GW is illustrated. Different color rings represent achievable range of different SF from GW and different color circles represent nodes with selected SF. Most of the end devices which close to the GW will fall into lowest SF (SF7) area section. End devices which close to the GW will probably select lowest SF all the time. Which results a lot of collisions between same SF transmissions. Collision number increases while end device number close to the GW increases.

\begin{figure}
\centering
\includegraphics[width=\linewidth]{collision}
\caption{Collision between nodes close to the GW.}
\label{fig:collision}
\end{figure}

@@

\section{Proposed Technique} \label{Proposed Technique}
\par The collision issue illustrated in Figure \ref{fig:collision} can be solved if the GW force some of the close end nodes to select higher SF even if they can able to communicate with lower SF. This may result lower collisions due to the orthogonality of different SF as shown in Figure \ref{fig:collision_solution_single_gw}. Higher SF assigned nodes are drawn with bold circle border in Figure \ref{fig:collision_solution_single_gw}. However, how GW assign SF to which node becomes an another issue. Increasing a node's SF should be done carefully since higher SF means higher time on air and higher time on air is more prone to collision with other higher SF transmissions. In multiple GW scenarios, this approach may increase the collision number with nodes in other GW's range. Thus, extra care must be taken for nodes in intersection area of GWs illustrated in Figure \ref{fig:collision_solution_multi_gw}.

@@

\begin{figure}
\centering
\includegraphics[width=\linewidth]{collision_solution_single_gw}
\caption{Collision avoidance proposal between nodes close to the GW by using higher SF.}
\label{fig:collision_solution_single_gw}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=\linewidth]{collision_solution_multi_gw}
\caption{Collision avoidance proposal for intersecting GWs.}
\label{fig:collision_solution_multi_gw}
\end{figure}

\section{Simulation Environment and Results} \label{Simulation Environment and Results}
\par A discrete event simulator is implemented in Python to study LoRa network performance considering SF orthogonality. The LoRaWAN SF simulation tool is publicly available at https://github.com/tugrulyatagan/simlorafs. Simulation tool supports custom LoRaWAN topologies as well as randomly generated LoRaWAN topologies. Simulator can generate uniformly distributed random topology with node number and gateway number input. Global simulation parameters are simulation duration (s), packet size (B), packet generation rate (p/s) and spreading factor assignment method. With these input, tool provide outputs of total generated packet number, successfully received packet number, interfered packet number, under sensitivity packet number, network packet delivery ratio (PDR), network throughput and total transmit energy consumption.

\subsection{Link Model}
\par Link quality of a wireless system can be expressed by the metric of link budget. Link budget is a measure of all gains and losses from transmitter to receiver. Link budget of a wireless link can be calculated as \cite{AN1200.22}:

@@

\par Where, $P^{dBm}_{RX}$ is the expected receive power at the receiver. $P^{dBm}_{TX}$ is the transmit power of the transmitter. $G^{dB}_{SYS}$ is the system gains such as transmitter and receiver antenna gains. $L^{dB}_{SYS}$ is the system losses such as transmitter and receiver line, circuit, antenna losses. $L^{dB}_{PATH}$ is the propagation path loss between transmitter and receiver antennas in open space. In our simulation we assume that total of system gains $G^{dB}_{SYS}$ and system losses $L^{dB}_{SYS}$ is +7 dBm.

\begin{table}
\centering
\caption{EU863-870 ISM Band LoRa Default Channels \cite{lorawan.regional.parameters}}
\label{table:max_tx_power}
\begin{tabular}{|c|c|c|c|}
\hline
\textbf{Channel} & \textbf{Frequency (MHz)} & \textbf{Bandwidth (kHz)} & \textbf{Max EIRP (dBm)} \\ \hline
      1 &     868.1 &   125 &   14 \\ \hline
      2 &     868.3 &   125 &   14 \\ \hline
      3 &     868.5 &   125 &   14 \\

\par Maximum transmit power for European ISM band LoRa default channels can be found in Table \ref{table:max_tx_power}. In our simulator we assume that nodes are always uses maximum allowed transmit power which is 14 dBm. Different channel transmissions are independent from each other. But, in this work we focuse on SF orthogonality. Thus, we only utilized single channel transmissions in our simulator.

\begin{table}
\centering
\caption{Gateway Sensitivity for different SFs \cite{SX1276}}
\label{table:gw_sf_sensitivity}
\begin{tabular}{|c|c|c|c|c|c|c|c|}
\hline
\multicolumn{2}{|c|}{\multirow{2}{*}{}} & \multicolumn{6}{c|}{\textbf{SF}} \\ \cline{3-8}
\multicolumn{2}{|c|}{}                  &    7 &    8 &    9 &   10 &   11 &   12 \\ \hline
\multirow{3}{*}{\textbf{BW (kHz)}}  & 125 & -123 & -126 & -129 & -132 & -133 & -136 \\ \cline{2-8}
                                    & 250 & -120 & -123 & -125 & -128 & -130 & -133 \\ \cline{2-8}
                                    & 500 & -116 & -119 & -122 & -125 & -128 & -130 \\ \hline
\end{tabular}
\end{table}

\par Receive sensitivity of a LoRa gateway for different SFs and BWs in dBm can be found in Table \ref{table:gw_sf_sensitivity} \cite{SX1276}. In our simulation, we use 125 kHz bandwidth which is also the only allowed bandwidth in EU863-870 regulation.

\par Free space propagation loss can be calculated as \cite{TR136.942}:

@@

\par Where $h$ is the gateway altitude and $f$ is the frequency of the signal. We assume that $h$ = 15 m and $f$ = 868 MHz. With these assumptions, propagation loss calculation become \cite{7996384}:

@@

\par If the received signal power is higher than the gateway sensitivity, then signal can be decoded by the receiver successfully when there is no interfering transmission at the transmission period.

\subsection{Interference Model}
\par We invite readers to refer to \cite{7996384}

@@

\par Transmission duration of a packet can be calculated by data rate $R_{b}$ and packet size $PS$. Data rate of a LoRa transmission is already expressed in Equation \ref{eq:bit_rate_sf}. Duration of a LoRa transmission:

\begin{equation} \label{eq:transmission_duration}
t_{x}|_{s} = \dfrac{8 \times PS|_{B}}{R_{b}|_{bps}}
\end{equation}

\begin{equation} \label{eq:p_interference}
P_{rc,y}^{interf} = \dfrac{P_{rc,y}(t_{interf})}{t_{x}}
\end{equation}

\begin{figure}
\centering
\includegraphics[width=\linewidth]{sf_pdr_r3000_g1_p001_s3600}
\caption{PDR of different SFs.}
\label{fig:sf_pdr}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=\linewidth]{gw_pdr_r3000_p001_s3600}
\caption{PDR of different gateway numbers.}
\label{fig:gw_pdr}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=\linewidth]{r_pdr_g1_p001_s3600}
\caption{PDR of different topology radius.}
\label{fig:r_pdr}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=\linewidth]{pr_pdr_r3000_g1_s3600}
\caption{PDR of different packet rates.}
\label{fig:pr_pdr}
\end{figure}

\begin{figure}
\centering
\includegraphics[width=\linewidth]{predict_pdr_r5000_g3_p001_s3600}
\caption{PDR of different prediction SF assignment scheme.}
\label{fig:predict_pdr}
\end{figure}

@@

\section{Conclusion} \label{Conclusion}
\par In this paper, after brief introduction about LPWAN technologies, we provide background information about one of the most promising LPWAN technology, LoRa. Then, we discuss about LoRa modulation basics and spreading factor assignment issue. We have implemented a discrete event simulator from scratch to study network performance of LoRaWAN and evaluate different SF assignment schemes. We have shown how same SF transmission collisions can be avoided and we proposed a machine learning based solution for collision with same SF transmissions. We illustrate simulation results for proposed machine learning SF assignment technique. Our simulation results show that, proposed technique can increase network performance for dense LoRaWAN network deployed close to GWs.

\par As for future works, transmit power optimization can be included to proposed machine learning scheme. In this paper, we assume nodes always use maximum transmit power for uplink transmission, however nodes close to the GW can decrease transmit power to save energy. This will make transmission more vulnerable to interference thus requires extra care. Optimizing another node parameter alongside SF can be interesting research. Also other machine learning methods can be investigated for SF optimization. Reinforcement learning can be a good candidate for this network optimization issue.

%% \par \cite{7815384} \cite{7803607} \cite{7996384} \cite{8090518} \cite{s17061193} \cite{8267219} \cite{8430542} \cite{8319183} \cite{8480649} \cite{AN1200.22} \cite{Bor:2016:LLW:2988287.2989163} \cite{8406255} \cite{DBLP:journals/corr/abs-1802-10338} \cite{finnegan2018comparative}
